---
title: "Final Exam"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

Please read all of the questions and tables carefully and follow all instructions. Each question has an allotted point value. Be as thorough and specific as possible; extra calculations and incorrect information, even in the presence of correct information, will result in point deductions. Be sure to show all formulas and all necessary work to answer the questions. You may upload your completed exam to the Elms course site.

::: callout-note
While this is an open-note exam, you are not to receive assistance from anyone else. As usual, the Honor Code applies.
:::

**Total points: 50**

## Conceptual Questions

Please include all work (and computations) necessary to answer the questions.

**Total points: 20 (2 points each)**

### Question 1

::: {.callout-tip}
2/2
:::

Suppose you estimate a bivariate regression model with a total sample size of 50 and obtain a standard error for $\beta_1$ of 2.50. What is the smallest regression coefficient ($\beta_1$) that you could have and still be able to reject the null hypothesis (of no relationship between $X_1$ and $Y$) at the 0.05 level with a one-tailed test?

```{r}
alpha <- 0.05
df <- 48

critical_value <- qt(1 - alpha, df)

standard_error <- 2.50

smallest_beta1 <- critical_value * standard_error
smallest_beta1
```

Based on the math above, we can determine that the smallest regression coefficient that you can have and still be able to reject the null hypothesis at the 0.05 level with a one-tailed test is 4.19306.

### Question 2

::: {.callout-tip}
2/2
:::

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and find that $\beta_1 = 4.20$, $SE_{\beta_1} = 1.90$, $t = 2.21$, $p = 0.01$, and a 95% confidence interval is \[0.48, 7.92\]. What is my best estimate of the effect of a two-unit change in $X$ on the mean of $Y$ in the population?

(4.20\*2)

Based on the math above, we can determine that the best estimate of the effect of a two-unit change in X on the mean of Y in the population is 8.40.

### Question 3

::: {.callout-tip}
2/2
:::

I conduct an OLS regression with a sample size of 90 and 5 independent variables. To determine a p-value for each coefficient, I would examine a t-distribution with how many degrees of freedom?

(90 - 5 - 1)

To determine the degrees of freedom for the t-distribution we must subtract the independent variables from the sample size, then take one away. Thus we would examine a t-distribution with 84 degrees of freedom to determine the p-value for each coefficient.

### Question 4

::: {.callout-tip}
0/2
:::

Suppose I regress $Y$ on $X$ and compute the mean response for $Y$ at some specified value of $X$. When determining the confidence interval around this mean response, which of the following will **NOT** have any effect on the width (or, size) of that confidence interval?

A.  Total sample size,

B.  Mean-squared error,

C.  The specified value of $X$,

D.  **None of the above: all of these (above) will affect the confidence interval**,

E.  There is not enough information to answer this question.

Based on the information above, we can determine the C. The specified value of $X$ is the correct answer.

The confidence interval (CI) for the mean response in regression depends on the standard error (SE), which is influenced by the mean-squared error and the total sample size. The calculation of the CI for the mean response doesn't involve the use of X. Although the precision of the estimate at a specified X can be affected by that specific value, it doesn't directly impact the CI for the mean response. The CI width is more broadly influenced by factors like SE, mean-squared error, and total sample size.

### Question 5

::: {.callout-tip}
2/2
:::

I regress $Y$ on $X$ and find that $\beta_1$ has a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  The lower and upper bound of a 95% confidence interval around $\beta_1$ will have the same sign,

B.  The absolute value for the t-statistic for $\beta_1$ will be greater than 1.96,

C.  A 90% confidence interval around $\beta_1$ will not contain zero,

D.  All of the above,

E.  None of the above and/or there is not enough information.

Based on the information above, we can determine that D. All of the above, is the most accurate statement.

### Question 6

::: {.callout-tip}
0/2
:::

I regress $Y$ on three independent variables -- $X_1$, $X_2$, and $X_3$ -- and I find the following 95% confidence intervals -- $\beta_1$: \[0.12, 1.45\], $\beta_2$: \[-0.01, 0.15\], and $\beta_3$: \[-0.64, -0.01\]. Which of the following statements is most accurate?

A.  $\beta_2$ and $\beta_3$ will have negative coefficients, and $\beta_1$ is statistically significant at the 0.05 level (two-tailed),

B.  $\beta_1$ will have a positive coefficient, and $\beta_1$ is the only statistically significant coefficient (of the three coefficients) at the 0.05 level (two-tailed),

C.  $\beta_1$ will have a positive coefficient, $\beta_2$ and $\beta_3$ will have negative coefficients, and only $\beta_1$ and $\beta_3$ are statistically significant at the 0.05 level (two-tailed),

D.  All of the above,

E.  **None of the above and/or there is not enough information.**

Based on the information above, C. $\beta_1$ will have positive coefficient, $\beta_2$ and $\beta_3$ will have negative coefficients, and only $\beta_1$ and $\beta_3$ are statistically significant at the 0.05 level is the most accurate statement.

B2 and B3 are deemed not statistically significant as their confidence intervals don't include zero. However, B1 is statistically significant since its confidence interval excludes zero, and it will be positive for the same reason.

### Question 7

::: {.callout-tip}
2/2
:::

Suppose I estimate a regression with two independent variables and obtain a $R^2$ of 0.40 where the Residual Sum of Squares is equal to 5,150. What does the Total Sum of Squares equal in this regression model?

(1-0.40) (5150/0.6)

Based on the math above, in this regression model the total sum of squares equals 8583.333.

### Question 8

::: {.callout-tip}
0/2
:::

I regress $Y$ on $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$). I find that $\beta_1 = -0.87$ with a two-tailed p value of 0.001 and $\beta_2 = 3.52$ with a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  $X_2$ has a larger substantive effect on $Y$ than $X_1$, but the effect of $X_1$ is more statistically significant,

B.  $X_1$ has a larger substantive effect on $Y$ than $X_2$ and $X_1$ is more statistically significant than $X_2$,

C.  $X_1$ has a larger substantive effect on $Y$ than $X_2$, but the effect of $X_2$ is more statistically significant,

D.  **None of the above and/or there is not enough information.**

Based on the information above, we can determine that A. $X_2$ has a larger substantive effect on $Y$ than $X_1$, but the effect of $X_1$ is more statistically significant is the most accurate statement.

### Question 9

::: {.callout-tip}
2/2
:::

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and obtain a $R^2$ of 0.45. Then, I regress $Y$ on both $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$) and obtain a $R^2$ of 0.53. Which of the following statements must be true?

A.  The correlation between $X_1$ and $Y$ is stronger than the correlation between $X_2$ and $Y$,

B.  The coefficient on $\beta_1$ is statistically significant in both models,

C.  The model with $X_1$ and $X_2$ explains more of the variation in $Y$ than a model with just $X_1$,

D.  All of the above: all of these statements are true.

Based on the information provided, C. the model with $X_1$ and $X_2$ explains more of the variation in $Y$ than a model with just $X_1$ is the correct answer.

Having only $X_1$ results in a lower R-squared, but incorporating $X_2$ leads to a higher R-squared, thus explaining more variability in the variable.

### Question 10

::: {.callout-tip}
1/2
:::

Based on the following regression equation, might the negative coefficient on the South variable be the result of southern states having a lower proportion of high school graduates than non-southern states? In no more than one sentence, explain your answer.

$$
Turnout = 34 + 0.5(Percent\ High\ School\ Graduates) – 5.9(South)
$$

The negative coefficient on the South variable is not really showing southern states having a lower proportion of high school graduates than non-southern states, as the coefficient displays the relationship between southern states and voter turnout.

::: {.callout-tip}
This is not really clear. What relationship is it showing? How do we interpret binary variables? 
:::

## Applied Questions

All datasets referenced below are posted on the ELMs course site.

```{r}
library(poliscidata)
library(readr)
library(tidyverse)
library(knitr)
library(broom)
library(modelsummary)
library(stargazer)
library(marginaleffects)
library(plotly)


world <- read_csv("world.csv")
```

### Question 1

*Points: 15*

Use the `world` dataset to answer the following questions. The relevant variables and their coding information are as follows: `literacy` indicates a country's literacy rate; `dem_score14` represents a country's level of democratization (higher values indicate greater democratization); `spendeduc` reflects the amount of public expenditures on education as a percentage of GDP; `gdp_10_thou` represents GDP per capita (one unit is \$10,000 USD); `educ_quality` indicates the average quality rating of a country's educational system; and `ungr9095` represents the percent average annual population growth.

#### Part A

*Points: 4*

::: {.callout-tip}
2/4
:::

Evaluate the bivariate relationship (using R) between a country's support for public education and its literacy rate. Report the results in a professionally formatted table. Do countries that spend more money on education appear to have a higher literacy rate?

```{r}
model_world_1 <- lm(literacy ~ spendeduc, data = world)
summary(model_world_1)

tidy_model_world_1 <- tidy(model_world_1)
tidy_model_world_1

world_titles <- c("Education Spending Percentage", "Intercept")
stargazer(model_world_1,
          covariate.labels = world_titles,
          dep.var.labels = c("Literacy Rate"),
          ci = TRUE,
          ci.level = 0.95,
          type = "text")
```

Based on the information, we can conclude that a one-unit increase in the Education Spending Percentage is estimated to result in a 1.721968 unit increase in a country's literacy rate. 

::: {.callout-tip}
... on average and holding all else constant. 
:::

This indicates a positive relationship between a country's support for public education and its literacy rate. The R\^2 value of 0.027 suggests that 2.7 percent of the variability in a country's literacy rate can be explained by its education spending percentage. There is a statistically significant relationship at the 0.1 level. Overall, it is evident that countries investing more in public education tend to display higher literacy rates.

::: {.callout-tip}
Avoid using "one-unit increase". To interpret these effects, you need to talk in terms of the specific variable units. 
:::

#### Part B

*Points: 5*

::: {.callout-tip}
4/5
:::

Estimate a second regression model (using R) that includes both a country's public expenditures on education and its level of democratization as independent variables. When controlling for democratization, does education expenditures exhibit a significant impact on literacy? Report the results in a professionally formatted table. Be sure to discuss the extent to which the results change, and if so, why they changed.

```{r}
model_world_2 <- lm(literacy ~ spendeduc + dem_score14, data = world)
summary(model_world_2)

tidy_model_world_2 <- tidy(model_world_2)
tidy_model_world_2

world_titles_2 <- c("Education Spending Percentage", "Level of Democratization")
stargazer(model_world_2,
          covariate.labels = world_titles_2,
          dep.var.labels = c("Literacy Rate"),
          ci = TRUE,
          ci.level = 0.95,
          type = "text")

modelsummary(model_world_2,
             coef_rename = c(spendeduc = "Public Expenditures on Education Percentage",
             dem_score14 = "Level of Country's Democratization"),
statistic = c("t-value = {statistic}", "Standard Error = {std.error}", "conf.int", "p.value"))
```

Based on the information above we cannot determine that spending on public education increases a country's literacy rate. This is because the p-value of 0.748 is much greater than the value of 0.05. Thus, when controlling for democratization rates, there is no statistical relationship between public education spending and a country's literacy rate.

::: {.callout-tip}
Be careful with your language when interpreting regression results. There is a statistical relationship: it is just not significant. Also, we can never determine a causal relationship using regression.
:::

#### Part C

*Points: 6*

::: {.callout-tip}
5/6
:::

Now estimate a third model (using R) that includes all of the variables listed above. Report the regression results in a professionally formatted table and interpret each coefficient. When interpreting the impact of democratization using this regression model, present/utilize a visual representation (using R) of the expected change in literacy as a function of democratization.

```{r}
model_world_3 <- lm(literacy ~ dem_score14 + spendeduc + gdp_10_thou + educ_quality + ungr9095, data = world)
summary(model_world_3)

tidy_model_world_3 <- tidy(model_world_3)
tidy_model_world_3

modelsummary(model_world_3,
             coef_rename = c( dem_score14 = "Level of Country's Democratization",
                              spendeduc = "Public Expenditures on Education Percentage",
                              gdp_10_thou = "GDP per Capita",
                              educ_quality = "Average Quality Rating of Country's Education",
                              ungr9095 = "Average Annual Population Growth Percentage"),
             statistic = c("t-value = {statistic}", "Standard Error = {std.error}", "conf.int", "p.value"))

plot_predictions(model_world_3, condition = "dem_score14")
```

Based on the provided information, we can interpret that the intercept being 55.9103 displays the value expected for a country's literacy rate when all independent variables are equal to zero.

::: {.callout-tip}
Again, you need to talk in terms of the variable units. Also, don't forget that these effects are on average and holding all else constant. 
:::

The coefficient for `dem_score14` is estimated at 2.5456, indicating that, on average, an increase of one unit in the democracy level of a country is associated with a 2.5456 unit increase. This relationship is also statistically significant.

The coefficient for `spendeduc` is estimated at 0.1124, suggesting that, on average, there is not a statistically significant relationship present between education spending and a country's literacy rate.

Moving on to the `gdp_10_thou` coefficient, which is estimated at 2.3598, we can infer that, on average, a one-unit increase in GDP per capita is associated with a 2.3598 unit increase in literacy rates. However, there is no statistically significant relationship observed.

The coefficient for `educ_quality` is estimated at 3.8118, indicating that, on average a one-unit increase in the average quality rating of a country's educational system is associated with a 3.8118 unit increase in the literacy rate. The model shows a statistically significant relationship at the 0.1 level.

Finally, the coefficient for `ungr9095` reveals that, on average a one-unit increase in the percent average annual population growth is associated with a decrease of -3.3259 units in the literacy rate of a country. Furthermore, there is a statistically significant relationship observed at the 0.001 level.

### Question 2

*Points: 15*

Using the `states` dataset, use R to regress the variable `obama08` (the percentage of a state's vote that President Obama received in the 2008 U.S. presidential election) on the following independent variables: `cig_tax` represents the amount of a state's cigarette tax (in dollars); `college` is the percentage of a state's population that graduated college (0 to 100); `union07` is the percentage of a state's workers that are union members (0 to 100); and `south` is a dichotomous variable coded as `1` if the state is located in the South (`0` otherwise). Report the results in a professionally formatted table. Answer the following questions.

```{r}
model_states <- lm(obama08 ~ cig_tax + college + union07 + south, data = states)

model_states$coefficients["south"] <- model_states$coefficients["southSouth"]
model_states$coefficients <- model_states$coefficients[!grepl("southSouth", names(model_states$coefficients))]

summary(model_states)

tidy_model_states <- tidy(model_states)
tidy_model_states

modelsummary(model_states,
             coef_rename = c(cig_tax = "State Cigarette Tax",
                             college = "Percentage of State College Graduates",
                             union07 = "Percentage of State Workers in a Union",
                             south = "State in the South"),
             statistic = c("t-value = {statistic}", "Standard Error = {std.error}", "conf.int", "p.value"))
```

#### Part A

*Points: 6*

::: {.callout-tip}
4/6
:::

Interpret the substantive results from this regression model. Be sure to discuss the direction, magnitude, and statistical significance of each slope coefficient. Does the `union07` variably exhibit a substantively significant impact on voting behavior in the 2008 election?

Based on the information provided by the regression model, we can analyze the coefficients as follows:

The coefficient for `cig_tax` is estimated at 0.5230, and the associated p-value of 0.815676 indicates that on average there is no statistically significant relationship. In a positive direction, we determine that the amount of a state's cigarette tax (in dollars) does not significantly impact the percentage of the state's vote that President Obama received in the 2008 U.S. presidential election.

::: {.callout-tip}
You needed to provide more information about the estimated effects of these variables in terms of their specific units. 
:::

Moving on to the `college` coefficient, which is estimated at 0.9646, the low p-value of 0.000299 suggests statistical significance (less than 0.001). With a positive slope direction, we can also indicate that on average a one-unit increase in the percentage of a state's population that graduated college corresponds to a 0.9646 percent increase in the state's vote for Obama.

The coefficient for `union07` is estimated at 0.7514, with a p-value of 0.003511, achieving statistical significance at the 0.01 level. On average, the positive slope direction implies that each one-unit increase in the percentage of a state's workers who are union members is associated with an increase of 0.7514 percentage points in the Obama vote for a state. Consequently, union membership in a state exhibits a substantively significant impact on voting behavior in the 2008 election.

Lastly, the `south` coefficient has an estimate of 1.3928, but with a p-value of 0.593179, it is not statistically significant. Despite the positive slope direction, we can determine that, on average, being located in the South does not have a significant impact on the percentage of a state's vote that Obama received in 2008.

#### Part B

*Points: 2*

::: {.callout-tip}
2/2
:::

Interpret the intercept coefficient. What does this represent and is it meaningful to interpret this coefficient on its own? Why or why not?

Analyzing the regression table, the intercept coefficient is 16.1280. The intercept signifies the estimated value when all independent variables are equal to zero. In this context, it suggests that the percentage of a state's vote that President Obama received in the 2008 U.S. presidential election, with all independent variables at zero, is estimated at 16.1280 units.

However, interpreting this coefficient on its own lacks meaningful context. This is due to the unrealistic expectation that all independent variables would simultaneously have a presence of zero. For a comprehensive understanding, it is essential to consider the interplay and impact of the other coefficients provided in the regression analysis.

#### Part C

*Points: 2*

::: {.callout-tip}
1/2
:::

What is the expected mean percentage of Obama's vote total in a Southern state with a cigarette tax of \$2.00, a state population where 30% of people are college graduates, and where 15% of workers in a state are union members?

16.1280 + (0.5230 \* *2.00) + (0.9646 \** 0.30) + (0.7514 \* *0.15) + (1.3928 \** 1)\

Based on the math above from the regression table, we can determine that 18.96889 is the expected mean percentage of Obama's vote total.

::: {.callout-tip}
The answer is 58.78%. Your `union` value should be 30, not 0.3.  
:::

#### Part D

*Points: 3*

::: {.callout-tip}
3/3
:::

What is a 90% confidence interval around the coefficient for the `south` variable.

```{r}
conf_interval_south <- confint(model_states, "south", level = 0.90)
conf_interval_south
```

Based on the information above, we can conclude that a 90 percent confidence interval around the coefficient for the `south` variable is (-2.954374, 5.739952).

#### Part E

*Points: 3*

::: {.callout-tip}
3/3
:::

What is the null hypothesis that the F-test in this regression output is testing?

The null hypothesis being tested by the F-test is that all of the coefficient independent variables in this model are equal to 0. With a F-statistic of 12.19 on 4 and 45 degrees of freedom and a p-value of 8.54e-07, we reject the null hypothesis (H0: All coefficients = 0). This suggests a significant linear relationship, indicating at least one predictor is statistically significant in explaining the variation in the dependent variable (Obama08).
