---
title: "Problem Set 9"
subtitle: "Due date: 4 December"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 40**

## Question 1

*Points: 10*

::: {.callout-tip}
10/10
:::

Table 1 below reports the results from two regression models. In Model 1, in Table 1, $Y$ is regressed on $X_1$ and, in Model 2, $Y$ is regressed on both $X_1$ and $X_2$. Why might $X_1$ be statistically significant at conventional levels in Model 1 but statistically insignificant in Model 2? Be as specific as possible.

Based on the table above we can claim several reasons as to why X1 is statistically significant at conventional levels in Model 1 but statistically insignificant in Model 2. First, the relationship between X1 and Y may be caused by the lurking variable Z, if we control for such variable we will not see a statistically significant relationship in Model 1. Next, the sample size may be at play. A small sample size may cause other variables to increase the standard error thus impacting the coefficient estimates. Finally, if the effect of X1 on Y alters across the values of X2 then we can determine the presence of an interactive effect on X1 and X2.

## Question 2

*Points: 10*

::: {.callout-tip}
7/10
:::

Using the `censusAggregate` dataset (posted on ELMs) --- which is survey data aggregated to the state level (1972-2000) --- estimate a regression with `vote` as the dependent variable and the following independent variables: `nonSouth`, `edr`, and `pcthsg`. Report the results in a professionally formatted table and interpret the regression results.

Also, create a figure to display the predicted values (with confidence intervals) for the effect of `pcthsg` on the turnout rate. Lastly, is it meaningful to interpret the constant term on its own? Why, or why not?

```{r}
library(tidyverse)
library(ggplot2)
library(modelsummary)

census_data <- read.csv("censusAggregate.csv")

model <- lm(vote ~ nonSouth + edr + pcthsg, data = census_data)
summary(model)

ggplot(census_data, aes(x = pcthsg, y = vote)) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Predicted Values with Confidence Intervals",
       x = "Percentage of High School Graduates",
       y = "Turnout Rate")
```

Based on the regression table above we can see that the model is statistically significant overall. The variables `nonSouth`, `edr`, and `pcthsg` all have a statistically significant relationship on the dependent variable `vote`. We can also see that the model shows the 28.13 percent variability in `vote` as the dependent variable.

::: {.callout-tip}
I need a more detailed interpretation of the effects of each variable on `vote`. 
:::

Interpreting the constant term on its own is not meaningful without the full context of the model. It is better to consider interpreting the entire model, where the effects of individual variables and their interactions can be included.

::: callout-note
`vote` is the turnout rate in a state in a given year (i.e., the number of people who voted divided by the number eligible to vote).

`nonSouth` is a dummy variable equal to `0` for Southern states and a `1` for non-Southern states.

`pcthsg` is the percentage of the population in a state that graduated high school.

`edr` is a dummy variable equal to `1` for states that used election-day registration and a `0` for states without election-day registration.
:::

## Question 3

*Points: 5*

::: {.callout-tip}
2/5
:::

Using the regression results from the previous question, evaluate the null hypothesis that the effects (i.e., regression coefficients) of `nonSouth` and `pcthsg` are jointly equal to zero. Can you reject the null hypothesis? Be sure to demonstrate how you reached a definitive conclusion.

Based on the regression results from the previous question, the p-value is much lower than the significance level of 0.05 and thus we can reject the null hypothesis that the variables `nonSouth` and `pcthsg` are jointly equal to zero.

Because of the statistically significant relationship between `nonSouth` and `pcthsg` in predicting the dependent variable, we can reach a definitive conclusion by examining the p-value. Because of how low the p-value is we can conclude that at least one of the coefficients is not equal to zero.

::: {.callout-tip}
You needed to conduct an f-test. 
:::

## Question 4

*Points: 15*

::: {.callout-tip}
15/15
:::

Using one of the other datasets available in the `poliscidata` package pick one dependent variable and two or more independent variables. Run a regression of the dependent variable on the independent variables. In your answer, describe why you picked the variables you did, produce a professionally formatted results table, and describe your statistical and substantive findings.

```{r}
library(poliscidata)
model <- lm(kerry04 ~ blkpct04 + union04, data = states)
summary(model)
```

For this question I chose to examine the relationship between percent voting for John Kerry in 2004 to the percent of black individuals in 2004 and percent of workers who are union members in 2004. I decided to examine this specific relationship because conventional wisdom is the Democratic candidates have a strong coalition of supporters consisting of black individuals and unions. I hypothesize that the higher presence of union members and black individuals in a state, the more support for John Kerry.

Based on the table above we can see that there is a statistically significant relationship between union workers in a state and support for John Kerry in 2004. However in this model, we do not see a statistically significant relationship between black individuals in a state and support for John Kerry. Overall, there is a statistically significant relationship present in the model and we can also determine there is a wide range of variability in the support for John Kerry in 2004.
