---
title: "Problem Set 8"
subtitle: "Due date: 20 November"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 30**

## Question 1

*Points: 5*

::: callout-tip
5/5
:::

For the following regression equation, $\hat{Y} = 8.5 + 6x + \epsilon$, the standard error for $\beta_0$ is 2.5, the standard error for $\beta_1$ is 3.5, and the sample size is 2000. Find the t-statistic, 95% confidence interval, and p-value (using a two-tailed test) for $\beta_1$.

Is $\beta_1$ statistically significant at the 0.05-level with a two-tailed test? Why or why not?

```{r}
beta_0_se <- 2.5
beta_1_se <- 3.5
n <- 2000

t_stat <- 6 / beta_1_se

df <- n - 2
t_crit <- qt(0.025, df)

lower <- 6 + t_crit * beta_1_se
upper <- 6 - t_crit * beta_1_se

p_value <- 2 * pt(-abs(t_stat), df)

cat("t-statistic: ", t_stat, "\n")
cat("95% confidence interval: [", lower, ", ", upper, "]\n", sep = "")
cat("p-value: ", p_value, "\n")
```

Based on the results above, we cannot reject the null hypothesis because the t-statistic of 1.714286 is less than the critical value of 1.96. Thus at the 0.05 level, beta_1 is statistically significant.

## Question 2

*Points: 5*

::: callout-tip
5/5
:::

Suppose you estimate an OLS regression and retrieve a $R^2$ value of 0.45. If the Total Sum of Squares (TSS) from that regression equals 4,700, what is the value for the Residual Sum of Squares (RSS)?

```{r}
RSS <- (4700)*(1 - 0.45)
RSS
```

Based on the math above, we can conclude the the value for the Residual Sum of Squares is 2585.

## Question 3

*Points: 5*

::: callout-tip
5/5
:::

Suppose you estimate a bivariate regression with a sample size of 102 and obtain a regression coefficient ($\beta_1$) of 5.0. What is the largest standard error that $\beta_1$ could have and still be statistically significant (i.e., reject the null hypothesis of no relationship) at the 0.05 level with a one-tailed test?

The standard error for a one-tailed test at the 0.05 level is 1646. SE(beta_1) = 5.0/1646 = 3.037

Based on the math above, we can conclude that the largest standard error that beta_1 could have and still be statistically significant is 3.037.

## Question 4

*Points: 5*

::: callout-tip
4/5
:::

Using the `states` dataset from the `poliscidata` package, produce a scatterplot of the variables `romney2012` and `hispanic10` (with `romney2012` as the dependent variable on the y-axis). Fit a regression line to the scatterplot. Describe the scatterplot and include a copy of it. Note any suspected outliers, if any (a visual inspection will suffice for this question).

```{r}
library(poliscidata)
data(states)

plot(romney2012 ~ hispanic10, data = states, main = "Scatterplot of Romney2012 and Hispanic10", xlab = "Hispanic10", ylab = "Romney2012")

abline(lm(romney2012 ~ hispanic10, data = states), col = "grey")
```

Based on the scatterplot above, we can see that there is a negative slope as the population of Hispanic individuals in a state increases the Romney share of votes decreases. The highest amount of Romney votes is present when the population of Hispanic individuals is less than 15 percent.

::: callout-tip
Do you suspect there to be any outliers?
:::

::: callout-note
The variable `romney2012` measures the percentage of the state's vote that Mitt Romney received in the 2012 presidential election, and `hispanic10` indicates the percentage of the state's population that identified as Hispanic in 2010.
:::

## Question 5

*Points: 10*

::: callout-tip
8/10
:::

Estimate a bivariate regression with `romney2012` as the dependent variable and `hispanic10` as the independent variable and report the results in a professionally formatted table. In as much detail as possible, describe (and interpret) the regression results.

```{r}
library(modelsummary)
library(stargazer)

bivariate_regression <- lm(romney2012 ~ hispanic10, data = states)

bivariate_regression

stargazer(bivariate_regression, type = "text")
summary(bivariate_regression)
```

::: callout-tip
Please make your coefficient labels meaningful for your reader.
:::

Based on the results shown above, we can conclude that the p-value for the population of Hispanic individuals is higher than the critical value.

::: callout-tip
This is a strange way to phrase this. Better to say whether or not this coefficient meets statistical significance. 
:::

The coefficients show a negative linear relationship as the hispanic10 is -0.1961 thus not statistically significant at a 0.001 and 0.05 because 0.182 as a p-value is much higher than the critical value. The r-squared value seen is 0.0368 is a low value and we can determine that there is a small difference between those who identify as Hispanic and voted for Romney. Finally, based on the t-value of -1.354 which is lower than the critical value we can determine that the value is not statistically significant at any of the significance levels.
